---
title: "flgaCBC"
author: "Burnett, JL"
date: "February 20, 2019"
output: pdf_document
---
```{r DEFINE.SOME.PARAMS}
startYear <- 1972
```


```{r libs, include=F}
# Install packages as necessary
list.of.packages <- c(
    "stringr","ggplot2",  "here",  "knitr", "tidyr", "reshape2", "readr", "dplyr", "devtools", "kableExtra", "stargazer","mgcViz" ,  "nlme", "mgcv", "gratia")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

# Load pkgs
lapply(list.of.packages, require, character.only = TRUE)

# Install git packages 
devtools::install_github("trashbirdecology/flgaCBC", force = F)
library(flgaCBC)
```

```{r setup, include = F}
# Set default chunk options
knitr::opts_chunk$set( warning = F, message = F,  fig.cap = "...", cache = T, eval=T, echo=F)
# Define directories and create as necessary
dataDir <- paste0(here::here(), "/data/")
```

```{r importMungeData}
## Import and munge Christmas Bird Count data & NACC species list
# Load CBC data (YYYY - 2015)
birds_1957to2015 <-
  read.csv(paste0(dataDir,"CBC_flga_1957to2015.csv"), stringsAsFactors = FALSE, check.names = F) %>% 
    gather(key = "key", value = "value", -(1:4), factor_key =T) %>% 
    filter(!value %in% c("CW", "cw")) %>% 
    mutate(value = as.integer(value))

# Load recent data (2016-2018)
birds2016_2018 <-  read.csv(paste0(dataDir,"CBC_flga_2016to2018_long.csv"), stringsAsFactors = FALSE, check.names = F) %>% 
    rename( key = species, 
            value = how_many)

# Merge orig data sets
birds <- full_join(birds_1957to2015, birds2016_2018) %>% 
    editNames() 

# Get NACC checklist (aou)
if(!exists("nacc")){nacc <- getNACC(remove.hypen = F)}


# See which species are in birds but not in nacc.list
missingSpp<-setdiff(unique(birds$key), nacc$key) %>% sort()
  ## If there are issues, I recommend fixing them in the function editNames.R, or just do it in this chunk..
        # Un resolved issues, if necessary
        # NACC does not contain the following species that occur in CBC data:
            # Gray-headed swamphen (Porphyrio poliocephalus)

# join NACC list with our CBC data
birds <- left_join(birds, nacc) 

# Get total species count (excludes groups (like Sp., or UNID))
birds <- estRichness(birds, missingSpp, ignore = missingSpp[c(9)] ) %>% 
    right_join(birds)

# Clear some temp objects
rm(nacc, birds_1957to2015, birds2016_2018)

```

# Introduction
CBC and BBS have become powerful tools for understanding large-scale patterns of population changes in North American birds, having been applied to guilds, and individual species.  
Restrictions and major biases, however, do apply to these citizen science data (Dunn et al. 2005). 
How well do trends in local CBCs mirror larger trends? Can we use insights from long-term local participants to identify possible causes of some of these trends? What have local birders observed that might help explain declines and increases. Habitat loss, succession, climate, dog fennel, prairie regrowth, drought, cyclic changes of drying, 
Are there processes other than actual population changes creating apparent population changes (superior birding skills (Sauer et al. 1994; Dunn et al. 2005), more knowledge of where birds might be hiding? More competition among groups? More bird feeder watchers (Dunn et al. 2005)? Better techniques such as playbacks of mobbing or screech owls? Stable populations but decreases caused by an increasing count effort that can only count a fixed number of individuals? Declining birding skills as we get older?) Changes in effort over time (Butcher and McCulloch 1990)

## On the CBC data

The count format (broken up into 11 teams and all day counts) was instituted in 1972. Therefore, we analyzed data from ß`r paste(startYear)` to `r paste(max(birds$year))`.

## Grouping of species

Let me know if you want any to change (see Table \@ref(tab:sppGroupTab)).

```{r groupSpecies}
# Get species groupings
sppGroups <- speciesGroups() 
# Make a data frame from output 
sppGroups <- stack(setNames(sppGroups, names(sppGroups)))

birds.new <- sppGroups %>% left_join(birds %>% select(family, key) %>% rename(values = key))

# see if there are discrepancies between sppGroups and birds common names
(to.edit.sppGroups <- birds.new %>% distinct(values, family) %>%  filter(is.na(family)))
    # the typos should now all just be spp that dont exist in your CBC data, rather than typos!
```

In some portions, we present results in groups of species according to various hypotheses. Species groupings are as follows: `r unique(sppGroups$ind)` (see Table \@ref(sppGroupTab)).


# Statistical analysis of population trends 

## Generalized additive models

We estimated species population trends in the Gainesville, Florida Christmas Bird Count (CBC; National Audubon Society REF) circle (FLGA) for the period of `r startYear` to 2015 using generalized additive models (GAMs, Hastie and Tibshirani 1990; Wood 2006). GAMs are a flexible implementation of generalized linear models in cases where species’ populations exhibit non-linear trends. GAMs optimize the predictability of the relationship between the response and predictor variable(s) while accounting for the noise associated with year-to-year fluctuations in species counts. 

We ran two types of analyses: (i) species-level trends and (ii) group-level trends.

### Model type 1: Species-level trends (I do not think we should use these models)

This model includes a factor for species identity. A smooth is produced for each species over time, and is also included as a fixed effect. We assumed the response variable distribution as Poisson. This means that the expected value after log transformation is modelled to be linearly dependent on the explanatory variables.


\begin{equation} 
g(E(y_g)) = \beta_0 f(year, species) + species + hours + \epsilon_i, \\
y_g ~ quasi-poisson \\
where g is species group
\end{equation} 

### Model type 2: Group trends 

We analyzed individual species’ trends using the following model:

\begin{equation}
g(E(y_species)) = \beta_0 f(year, species) + hours + \epsilon_i, \\
y_species ~ quasi-poisson
where g is species group
\end{equation}


# Results

- Unless we plan on expicitly adding the species counts across groups, we should just interpret individual species' trends in groups. 
- Many species trends were non-linear over the study period. GAM optimizes the residuals by fitting knots in the model (count, response value), so if the trend is linear, it will fit it.
    * However, I would like to go back through and run *glm(bpph~year)* or *glm(count~year)* on those. 
    * Another issue is species which do not occur or left the area during our study period. Right now our model assumes missing data as zero (e.g. see Figure \@ref(fig:hofiEx)). 
    * Another issue is species whose counts are highly variable (e.g., Red-Winged Blackbird; see Figure \@ref(fig:rwbbFig))
- Around 1980 the hours effort and number of participants increased linearally, therefore we did not fit a smooth to the hours effort (see Figure \@ref(fig:effortPlots)). 
- I also looked a temporal autocorrelation (see [the plots in this link]()). It's strange because the effect of temporal autocorrelation for many species was different at 1 & 3 yeras vs. 0 and 2 years
- We controlled for temporal autocorrelation by taking first, second, adn third differences (AR1, AR2, and AR3). We chose the best fitting model for each species by running likelihood ratios
<!-- - A GAM basically a GLM where the linear preditor(s) are influenced by some unknown smooth function, and a GAMM is a GLMM where "..." -->

## Analysis results
```{r runModels, eval = F}
# Add groups to the bird data
birds.grouped <- full_join(birds, birds.new %>% rename(key = values) %>%  na.omit(family)) %>%
    distinct()

# Run some basic GAMs.
Run the gam for species (ind by spp)
gam.results.ind <- runGams(df = birds.grouped, family = "quasipoisson",
                           group.var = "ind",
                           rds.filename =  "groupLevelWithFixed",
                           mod.formula  = as.formula("value ~ s(year, by = as.factor(key)) + key + hours"),
                           fig.out.path = NULL,
                           year.start = 1973,
                           zero.thresh = 0.25,
                           length.out = 200)
gam.results.ind     <-readRDS(paste0(here::here(),"/Results/gamResults_","groupLevelWithFixed",  Sys.Date(),".rds"))


# Run the gam for species (key)
gam.results.key <- runGams(df = birds.grouped, family = "quasipoisson",
                   group.var = "key",
                   rds.filename =  "speciesLevel",
                   mod.formula  = as.formula("value ~ s(year) + hours"),
                   fig.out.path = NULL,
                   year.start = 1973,
                   zero.thresh = 0.11,
                   length.out = 200)

gam.results.key     <-readRDS(paste0(here::here(),"/Results/gamResults_","speciesLevel",  Sys.Date(),".rds"))

```


```{r effortPlots, fig.cap= "Annual effort and species richness for the FLGA CBC circle. Dashed line indicates year we begin analyses.", include=T, echo = F }
hours.p <- ggplot(data = birds)+
    geom_line(aes(year, hours))+
    theme_bw()+
    theme(plot.margin=grid::unit(c(0,0,0,0), "mm"))+
    ylab("hours effort") +
    xlim(c(min(birds$year), max(birds$year)))+
    geom_vline(xintercept = startYear, linetype = 2, color = "red")
    
rich.p <- ggplot(data = birds %>% distinct(year, rich))+
    geom_line(aes(year, rich))+
    theme_bw() + 
    theme(plot.margin=grid::unit(c(0,0,0,0), "mm"))+
    ylab("# species")+
    xlim(c(min(birds$year), max(birds$year)))+
    geom_vline(xintercept = startYear, linetype = 2, color = "red")

partic.p <- ggplot(data = birds %>% distinct(year, participants))+
    geom_line(aes(year, participants))+
    theme_bw() + 
    theme(plot.margin=grid::unit(c(0,0,0,0), "mm"))+
    ylab("# participants")+
    xlim(c(min(birds$year), max(birds$year)))+
    geom_vline(xintercept = startYear, linetype = 2, color = "red")
    
gridExtra::grid.arrange(rich.p, hours.p, partic.p)
```

```{r sppGroupTab}
# Munge data frame for table
birds.new.wide <- birds.new %>% 
    filter(!is.na(family)) %>% 
    distinct(family, values, ind) %>%
    rename(species = values, 
           sppGroup = ind) %>% 
    dplyr::select(sppGroup, family, species) %>% 
    mutate(family = as.character(family)) %>% 
    arrange(sppGroup, family)

# row.ind <-  birds.new.wide %>% 
#     group_by(sppGroup) %>% 
#     summarise(row.ind = n()) %>% 
#     mutate(row.ind = cumsum(row.ind))
# row.ind <-  matrix(row.ind$row.ind, nrow=2) # to get every otehr row value for table styling ( custom gray rows)
# row.ind <- c(row.ind[1, 1], row.ind[2, 1],
#                   row.ind[1, 3], row.ind[2, 3],
#                   row.ind[1, 5], row.ind[2, 5])        
# Maake the table
birds.new.wide %>%
  kable(format    = "latex", 
      longtable = T, 
      booktabs  = T, 
      caption   = "Species groups for analysis") %>% 
    column_spec(1:2, bold = T) %>% # col_spec BEFORE collapse rows else weird shit happens.
    collapse_rows(columns = 1:2, 
                  latex_hline = "major",
                  valign = "middle") %>%
    kable_styling(full_width = F, font_size = 9)#, latex_options = c("repeat_header"))
```

```{r sppAnalysedTab}

if(!exists("birds.grouped")) birds.grouped <- full_join(birds, birds.new %>% rename(key = values) %>%  na.omit(family)) %>%
    distinct()

gam.results.key     <-readRDS(paste0(here::here(),"/Results/gamResults_","speciesLevel",  Sys.Date(),".rds"))
# Maake the table
t = unique(names(gam.results.key))
t %>% 
      kable(format    = "latex", 
      longtable = T, 
      booktabs  = T, 
      caption   = paste0("Species analysed: ", length(t))) %>% 
    kable_styling(full_width = F, font_size = 9)#, latex_options = c("repeat_header"))
```

```{r sppNotAnalysedTab}
# Maake the table
t=birds.grouped %>% distinct(key) %>% filter(!key %in% unique(names(gam.results.key)))
 t %>%
  kable(format    = "latex", 
      longtable = T, 
      booktabs  = T, 
      caption   = paste0("Species **NOT** analysed: ", length(t))) %>% 
    kable_styling(full_width = F, font_size = 9)#, latex_options = c("repeat_header"))
```

```{r hofiEx, fig.cap="Example of missing data prior to year YYYY. We might want to go back through and adjust our start year for those species that did not occur in the region before or after some year (**i.e. will need expert advice**)."}
ts= birds.grouped %>% filter(key == "House Finch") %>% 
    distinct(year, key, value, hours)
ggplot(ts, aes(year, value))+
    geom_point()+
    geom_smooth()+
    ylab("House Finch")
```

```{r rwbbFig, cache = T, include=T,echo=F, eval=T, fig.cap="HOFI example -- species entering or leaving region for sequential years within our study period."}

ts=birds.grouped %>% filter(key == "Red-Winged Blackbird") %>% 
    distinct(year, key, value, hours)
ggplot(ts, aes(year, value))+
    geom_point()+ylab("Red-Winged Blackbird")+
    # geom_smooth(method='lm')+
    geom_smooth(method ="gam")
```



<!-- # References not yet defined in .bib -->

<!-- Fewster, Rachel M., et al. "Analysis of population trends for farmland birds using generalized additive models." Ecology 81.7 (2000): 1970-1984. -->
<!-- Hastie, Trevor J., and Robert J. Tibshirani. Generalized additive models. Vol. 43. CRC Press, 1990. -->
<!-- R Core Team (2015). R: A language  and environment for statistical computing. R Founding for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/. -->
<!-- Wood, S.N. (2006) Generalized Additive Models: An Introduction with R. Chapman and Hall/CRC. -->
